---
title: "STAN Examples"
author: Songpeng Zu
date: "`r format(Sys.time(), '%d %B %Y')`"
urlcolor: blue
fontsize: 10pt
output:
  html_document
---

# Set STAN Environment

* [Install cmdstanr in R](https://mc-stan.org/cmdstanr/articles/cmdstanr.html)

```{r eval=FALSE}
install.packages("cmdstanr", repos = c("https://mc-stan.org/r-packages/", getOption("repos")))

## The two are used for STAN result analysis.
install.packages("bayesplot")
install.packages("posterior")
```
* Install cmdstan either from the [cmdstan repository](https://github.com/stan-dev/cmdstan/wiki/Getting-Started-with-CmdStan) or by [cmdstanr](https://mc-stan.org/cmdstanr/reference/install_cmdstan.html)


* Set the basic environment
```{r results = "hide", warning = FALSE, message = FALSE}
library(cmdstanr)
library(bayesplot)
library(posterior)

## for bayesplot plotting
color_scheme_set("brewer-Spectral")

## If you use cmdstanr install cmdstan, you can ignore this.
set_cmdstan_path(path = paste(Sys.getenv("HOME"),
  "softwares",
  "cmdstan-2.23.0", sep = "/"))

## set the parameter for STAN sampling with multi-core support.
options(mc.cores = 3)
```

# Example 1: Multivariate normal with missing data

## Data
 Our observations are from $\mathcal{N}(0, \Sigma)$, where $\Sigma$ is unknown.
 
\begin{pmatrix}
1 & -1 & -1 & 1 & 2 & 2 & -2 & -2 & * & * & * & *\\
1 & -1 & 1 & -1 & * & * & * & * & 2 & 2 & -2 & -2
\end{pmatrix}
  Each column corresponds to one sample, and $*$ are missing data.
  
## How we infer $\Sigma$ using STAN 

* load the model
```{r cache = TRUE}
camel_mdl <- cmdstan_model("camel/camel2.stan", compile = TRUE)
camel_mdl$print()
```

* sampling the parameter

```{r cache = TRUE, warning = FALSE, message = FALSE}
camel_sampler <- camel_mdl$sample(data = NULL,
  seed = 355113,
  parallel_chains = 4,
  show_message = TRUE,
  ## increase adapt delta, default is 0.80
  adapt_delta = 0.85,
  iter_sampling = 1000
  ## output_dir = "./"
)
camel_draws <- camel_sampler$draws()
```

* show the posterior results
```{r cache=TRUE, collapse = TRUE, fig.width = 20}
bayesplot::mcmc_combo(camel_draws,
  pars = c("rho"), widths = c(1, 3),
  combo = c("dens_overlay", "trace")
 )
```

# Example 2: Finite normal mixture model

## load STAN script for both simulation and inference
```{r cache = TRUE, warning = FALSE, messgae = FALSE}
mixture_model <- cmdstan_model("mixture/normal_mixture_k_order.stan",
  compile = TRUE
)
```

## Generate the simulation data
```{r cache = TRUE, warning = FALSE, messgae = FALSE}
N <- 1200
N1 <- 400
N2 <- 400
musig <- 10.0
alpha <- c(1.0, 1.0, 1.0)
sigma <- c(1.0, 1.0, 1.0)

sim_sampler <- mixture_model$sample(
  data = list(
    K = 3,
    N = N,
    N1 = N1,
    N2 = N2,
    y = rep(1, N),
    musig = musig,
    alpha = alpha,
    sigma = sigma
  ),
  seed = 355113,
  chains = 1,
  iter_sampling = 1,
  parallel_chains = getOption("mc.cores", 1),
  show_message = TRUE,
  init = list(
    list(
      theta = c(N1 / N, N2 / N, 1 - N1 / N - N2 / N),
      mu = c(-5.0, 0.0, 5)
      ## sigma = c(0.5, 1, 1)
    )
  ),
  fixed_param = TRUE
)

sim_draws <- sim_sampler$draws()
## dimnames(sim_draws)$variable[1:10]

gy <- c(sim_draws[1, 1, grep("gy", dimnames(sim_draws)$variable)])
hist(gy, breaks = 50)
```


## Sampling with STAN

```{r cache = TRUE, warning = FALSE, messgae = FALSE}
sim_sampler <- mixture_model$sample(
  data = list(
    K = 3,
    N = N,
    N1 = N1,
    N2 = N2,
    y = gy,
    musig = musig,
    alpha = alpha,
    sigma = sigma
  ),
  seed = 355113,
  chains = 4,
  iter_sampling = 1000,
  fixed_param = FALSE,
  ## init = list(
  ##   list(
  ##     theta = c(0.4, 0.3, 0.3),
  ##     mu = c(-4.0, 0.0, 4),
  ##     sigma = c(10, 10, 10)
  ##   ),
  ##   list(
  ##     theta = c(0.2, 0.3, 0.5),
  ##     mu = c(-10.0, 0.0, 10),
  ##     sigma = c(10.0, 10, 10.0)
  ##   )
  ## ),
  parallel_chains = getOption("mc.cores", 1),
  ## output_dir = "./",
  show_message = TRUE
)
```
## View the results.
```{r cache = TRUE, warning = FALSE, message = FALSE, fig.width = 15}
ey <- sim_sampler$draws()
bayesplot::mcmc_combo(ey[, , grep("mu", dimnames(ey)$variable)],
  widths = c(1, 3),
  combo = c("dens_overlay", "trace"))
```

## If we set unordered $mu$ in STAN

```{r cache = TRUE, warning = FALSE, message = FALSE, include = FALSE, echo =FALSE}
umixture_model <- cmdstan_model("mixture/normal_mixture_k_unorder.stan",
  compile = TRUE
)

umixture_sampler <- umixture_model$sample(
  data = list(
    K = 3,
    N = N,
    N1 = N1,
    N2 = N2,
    y = gy,
    musig = musig,
    alpha = alpha,
    sigma = sigma
  ),
  seed = 355113,
  chains = 4,
  iter_sampling = 1000,
  fixed_param = FALSE,
  ## init = list(
  ##   list(
  ##     theta = c(0.4, 0.3, 0.3),
  ##     mu = c(-4.0, 0.0, 4),
  ##     sigma = c(10, 10, 10)
  ##   ),
  ##   list(
  ##     theta = c(0.2, 0.3, 0.5),
  ##     mu = c(-10.0, 0.0, 10),
  ##     sigma = c(10.0, 10, 10.0)
  ##   )
  ## ),
  parallel_chains = getOption("mc.cores", 1),
  ## output_dir = "./",
  show_message = TRUE
  )
```

```{r cache = TRUE, warning = FALSE, message = FALSE, fig.width = 15}
uy <- umixture_sampler$draws()
bayesplot::mcmc_combo(uy[, , grep("mu", dimnames(uy)$variable)],
                      widths = c(1, 3),
                      combo = c("dens_overlay", "trace"))

```
